\LARGE{\bf Parallelizing a Network Simulator}

### 2009-03-17 doc/paper.tex
git-svn-id: https://nocs.stanford.edu/cgi-bin/svn.cgi/booksim2.0/tags/pbooksim@1160 ce7aba0d-9f81-40a6-a495-cecae84b0515

\documentclass[twocolumn]{article}
\usepackage{amsmath}
\usepackage{epsfig}
\usepackage[all]{xy}
%\usepackage{multicol}
\usepackage{indentfirst}
\usepackage{algorithm}
\usepackage{algorithmic}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{9.5in}
\setlength{\topmargin}{0in}
\setlength{\oddsidemargin}{0in}
\setlength{\evensidemargin}{0in}
\setlength{\topmargin}{0in}
\setlength{\headheight}{0in}
\setlength{\headsep}{0in}
\setlength{\topskip}{0.5in}
%% be sure to change the:
% Section number, Day, and time
% Date and year
% Name
% spellcheck using ``aspell -c filename'' in Unix
%\begin{center}
%\LARGE{\bf Parallelizing a Network Simulator}
%% PUT IN YOUR SECTION, DAY, AND TIME
%%example:
%% Section 7 - Thursday 8:30 AM\\

@@

\subsubsection{Implementation}
Lock-free queues are based on the universal lock-free algorithm
\cite{LF} that assumes that any access to the shared data structure can
see the data structure in any potential partially updated state.  The
algorithm then allows for the access to help other accesses to
complete before actually completing itself.  While this may sound
conceptually simple, the combinatorial number of states of the data
structure and different readers and writers is astounding.\\
~\\
The lock-free queue implementation that we use begins by having a sentinel
at the head of the queue whose value is meaningless.  This eases some of
the implementation details of the queue.  Initially both the head and tail
pointers of the queue point to the sentinel.  Whenever a new item is added
to the queue, the tail pointer is moved to point to the end of the list.
Whenever an item is popped from the queue, the head is moved to point to
the new sentinel, the next item in the list.  In this case the old
sentinel can then be reclaimed.  Each element in the lock-free queue is
wrapped in an AtomicReference object.  This object ensures that the any
items in the queue can only be written to atomically.  A diagram of how
this works can be seen in Figure \ref{lfqueue}.\\
~\\
\begin{figure}[h]
\centering
\includegraphics[width=3in, height=2.25in]{lfqueue.eps}
\caption{This figure demonstrates the layout of the objects associated with a lock-free queue of length 2 (the sentinel value does not count to towards the length of the list).  Note that the pointer to the next element in the list, does not point directly to next list node, but instead to the wrapper AtomicReference class.  Also note that the sentinel is not an actual maintained pointer, but will always point to the same AtomicReference as the Head Node.  It is included in this diagram for clarity.}
\label{lfqueue}
\end{figure}

@@

\subsection{Analyzing Synchronization Mechanisms}
The first step in analyzing our parallelized network simulator was to
determine the performance of the different synchronization mechanisms that
we had employed at different points in development.  Figure \ref{synch}
shows the performance of the various synchronization mechanisms on a
single node of the Hotbox cluster that consists of two Intel Xeon Quad
cores.  We only ran experiments up to 8 threads as this is the maximum
number of hardware thread contexts available without encountering the
overhead of using ethernet to communicate between nodes.\\
\begin{figure}[h]
\centering
\includegraphics[width=3in,height=3in]{synch.eps}
\caption{\label{synch} This figure shows the performance of the different
synchronization mechanisms on up to eight threads on the Hotbox cluster.}
\end{figure}
~\\
From these results we can see that our hypotheses were confirmed regarding
the performance of the various synchronization mechanisms.  Barriering
every cycle added significant overhead to performance and ultimately
prevented even 8 threads from extracting much parallelism at all.
Surprisingly the same result was discovered when the barrier was removed and
all the queues were replaced with lock-free queues.  However, once we were
able to only implement the shared channels with lock-free queues, we could see a noticeable performance gain and even some
additional scalability.  Finally, once we fused several of the loops being
executed within the program, we were able to improve some of the temporal
locality of data use and thereby achieve a little additional
performance.\\
~\\
Despite our efforts however, we were only able to extract a two fold
speedup using eight threads, far below a linear speedup.  This was very
perplexing to us as we had consciously attempted to break up the work into
parallel computations that required very little inter-thread
communication.  In order to determine exactly what was causing us to lose
performance we then proceeded to run several experiments to find the
bottlenecks in the program.

@@

\includegraphics[height=3in,width=3in]{smp.eps}

@@

\includegraphics[height=3in,width=3in]{probsize.eps}

@@

\includegraphics[height=3in,width=3in]{niagara.eps}

>>>>

\documentclass[twocolumn]{article}
\usepackage{amsmath}
\usepackage{epsfig}
\usepackage[all]{xy}
%\usepackage{multicol}
\usepackage{indentfirst}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage[top=1in, bottom=1in, left=0.75in, right=0.75in]{geometry}
\setlength{\columnsep}{2pc}
\usepackage[small]{caption}
%% be sure to change the:
% Section number, Day, and time
% Date and year
% Name
% spellcheck using ``aspell -c filename'' in Unix
%\begin{center}
%\LARGE{\bf Parallelizing a Network Simulator}
%% PUT IN YOUR SECTION, DAY, AND TIME
%%example:
%% Section 7 - Thursday 8:30 AM\\

@@

\subsubsection{Implementation}
Lock-free queues are based on the universal lock-free algorithm
\cite{LF} that assumes that any access to the shared data structure can
see the data structure in any potential partially updated state.  The
algorithm then allows for the access to help other accesses to
complete before actually completing itself.  While this may sound
conceptually simple, the combinatorial number of states of the data
structure and different readers and writers is astounding.\\
~\\
The lock-free queue implementation that we use begins by having a sentinel
at the head of the queue whose value is meaningless.  This eases some of
the implementation details of the queue.  Initially both the head and tail
pointers of the queue point to the sentinel.  Whenever a new item is added
to the queue, the tail pointer is moved to point to the end of the list.
Whenever an item is popped from the queue, the head is moved to point to
the new sentinel, the next item in the list.  In this case the old
sentinel can then be reclaimed.  Each element in the lock-free queue is
wrapped in an AtomicReference object.  This object ensures that the any
items in the queue can only be written to atomically.  A diagram of how
this works can be seen in Figure \ref{lfqueue}.\\
~\\
\begin{figure}[h]
\centering
\includegraphics[width=3in]{lfqueue.eps}
\caption{This figure demonstrates the layout of the objects associated with a lock-free queue of length 2 (the sentinel value does not count to towards the length of the list).  Note that the pointer to the next element in the list, does not point directly to next list node, but instead to the wrapper AtomicReference class.  Also note that the sentinel is not an actual maintained pointer, but will always point to the same AtomicReference as the Head Node.  It is included in this diagram for clarity.}
\label{lfqueue}
\end{figure}

@@

\subsection{Analyzing Synchronization Mechanisms}
The first step in analyzing our parallelized network simulator was to
determine the performance of the different synchronization mechanisms that
we had employed at different points in development.  Figure \ref{synch}
shows the performance of the various synchronization mechanisms on a
single node of the Hotbox cluster that consists of two Intel Xeon Quad
cores.  We only ran experiments up to 8 threads as this is the maximum
number of hardware thread contexts available without encountering the
overhead of using ethernet to communicate between nodes.\\
\begin{figure}[h]
\centering
\includegraphics[width=3in]{synch.eps}
\caption{\label{synch} This figure shows the performance of the different
synchronization mechanisms on up to eight threads on the Hotbox cluster.}
\end{figure}
~\\
From these results we can see that our hypotheses were confirmed regarding
the performance of the various synchronization mechanisms.  Barriering
every cycle added significant overhead to performance and ultimately
prevented even 8 threads from extracting much parallelism at all.
Surprisingly the same result was discovered when the barrier was removed and
all the queues were replaced with lock-free queues.  However, once we were
able to only implement the shared channels with lock-free queues, we could see a noticeable performance gain and even some
additional scalability.  Finally, once we fused several of the loops being
executed within the program, we were able to improve some of the temporal
locality of data use and thereby achieve a little additional
performance.\\
~\\
Despite our efforts however, we were only able to extract a two fold
speedup using eight threads, far below a linear speedup.  This was very
perplexing to us as we had consciously attempted to break up the work into
parallel computations that required very little inter-thread
communication.  In order to determine exactly what was causing us to lose
performance we then proceeded to run several experiments to find the
bottlenecks in the program.

@@

\includegraphics[width=3in]{smp.eps}

@@

\includegraphics[width=3in]{probsize.eps}

@@

\includegraphics[width=3in]{niagara.eps}
